\begin{abstract}

Molten salt reactors (MSRs) are a class of nuclear reactor which uses a molten 
ionic liquid as either the coolant or also as the fuel. While a 8 MWth MSR was
successfully operated in the 1960s it was not until the early 2000s that MSRs
gained widespread attention. Since then MSRs have enjoyed plentiful research
support.
Despite such support only one
general MSR fuel cycle analysis tool is available for use by the research
community and even this this tool lacks features necessary to modelling an MSR
,as such possibly providing answers of a lower quality. 

In this work a method is proposed and implemented within the 
SERPENT 2 reactor physics Monte-Carlo code. This method, named ADER - the
Advanced Depletion Extension for Reprocessing - is a seamlessly incorporated
source code modification to the SERPENT 2 base code which allows the user
to define arbitrary collections of elements, isotopes, and chemicals as well
as relationships among them.

Furthermore ADER allows the user to specify a variety of mass flows subject to
the constraints as defined through the collections of elements, isotopes, and
chemicals the user has structured. Along with support for constraints involving
both corrosion and nuclear control concerns, ADER allows the user to optimize
the solution against a quantity of interest such as total uranium fed into the
system for example.

Through these structures
much of the complex chemistry, corrosion modelling, and nuclear concerns of
operating a MSR can be linearized and solved against an optimization target,
which is necessary given the unmatched number of constraints and variables in
such a problem space. Linearization reduces the problem complexity and
eliminates concerns over local versus global optimization targets. Within the
typically narrow operating parameters of MSRs linearization is an appropriate
approximation to the higher dimensional equations representing the phenomenon
involved.

This linear system and optimization target may then be
passed to a linear optimization
solver, in this work the CLP library as part of the COIN-OR package, from which
an optimized system material composition and material flows solution may be 
found. ADER then uses this solution to create a brand new depletion matrix 
which SERPENT 2 then solves using the CRAM approximation method. 

From this
algorithm a more accurate modelling of MSR fuel cycles and physics may be
arrived at through the consideration of chemistry driven limitations,
corrosion driven limitations, nuclear driven limitations, and operator driven
limitations. Results from this implemented method indicate that ADER drives
the MSR fuel cycle simulations towards a more physically representative
result. Unfortunately, as detailed later in this work, an underlying and
pernicious numerical instability issue was uncovered within the linear
optimization library selected for this work. Any future work on this method
must begin with the adoption of a quadruple-precision floating-point linear
optimization library over the current implementation of CLP as used in ADER.

In the following chapters an introduction to MSRs and their fuel cycle modelling
is given. Following this the theory behind ADER and its implementation within 
SERPENT2 is discussed after which the results from one of a few numerically
stable simulations is presented after which concluding remarks are given.
\end{abstract}
